Unscrambling arrested development: temporal ordering and registration of cross-sectional imaging data

### 2014-04-10 image_analysis_paper/pnas_draft.tex
more edits to paper

\dropcap{E}xperimental studies of developmental dynamics fall in two broadly defined categories: longitudinal and cross-sectional \cite{diggle2002analysis}.
%
In longitudinal studies, developmental progress is monitored over time for the same embryo.
%
In a cross-sectional study, one embryo contributes only a single snapshot of a chemical or morphological process along its developmental trajectory, and the developmental dynamics must be reconstructed from multiple snapshots of different embryos.
%
Both of these sampling schemes have their advantages and limitations, and both are extensively used by developmental biologists.
%
Here we focus on cross-sectional studies, which have a time-honored history and still present the only option for most organisms.
%
In a typical cross-sectional study, a group of developing embryos is fixed using a procedure that arrests their development and stained with chemicals that help visualize a handful of cellular processes.
%
Fixed embryos are then imaged using any given number of microscopy techniques.
%
Recent advances in large-scale physical manipulation and imaging of embryos have produced rapidly increasing volumes of cross-sectional data, in which every embryo is observed at a different geometric orientation and developmental time point.
%
Importantly, the ``age'' of any given embryo arrested in its development is not known to high accuracy.
%
In general, it is only known that a collection of embryos belongs to a certain time window.
%
In order to mine such image datasets for the developmental dynamics, snapshots of different embryos must be spatially aligned or registered to factor out the relevant symmetries (i.e., translations and rotations), and ordered in time.
%
We present a general computational framework which greatly accelerates -and can combine- both of these tasks.

Temporal ordering and registration of images is straightforward when the number of images is small and differences between them are visually apparent.
%
As an example, Figure~\ref{fig:fish} shows a caricature of fish development, combining the processes of growth and color patterning.
%
In this case, temporal ordering can be accomplished by arranging the fish by size, which uniquely correlates with the progress of development.
%
Image registration is based on obvious morphological landmarks, such as the position of head and fins.
%
On the other hand, real data poses nontrivial challenges, due to the number of images, measurement noise and variability, and the subtlety of the developmental changes.

To illustrate the issues associated with temporal ordering and registration of images, we use a data set from an imaging study of signaling protein expression in the early fruit fly embryo, one of the leading model systems for studies of developmental dynamics.
%
This dataset comprises $49$ fluorescence microscopy images of different embryos fixed during the $3^{rd}$ hour of their development.
%
\begin{figure}[t]
\includegraphics[width=8.4cm]{fig1}
\caption{Schematic illustration of reconstructing dynamics from cross-sectional data. (a) Fish, each in a different orientation and a different stage of development. (b) Fish, now rotationally registered and temporally ordered. For this example, the registration and ordering is easy to do ``by hand" because the data set is small and the developmental changes are easy to recognize.}
%\label{fig:fish}
\customlabel{fig:fish}{1}
\customlabel{subfig:fish_unordered}{\ref{fig:fish}a}
\customlabel{subfig:fish_ordered}{\ref{fig:fish}b}
\end{figure}
%
The fruit fly embryo is shaped like a rice grain, approximately $1/2$~mm long.
%
By end of the $2^{nd}$ hour of development, sequential divisions of a fertilized egg generate a system where $\sim 6,000$ nuclei are uniformly arranged under the common plasma membrane.
%
During the $3^{rd}$ hour, the embryo undergoes the process of cellularization, whereby membranes that enclose nuclei into different cells grow in thickness (Figure~\ref{subfig:membranes}).
%
Concurrently with these morphological changes, concentration profiles of multiple regulatory molecules form inside the embryo, subdividing it into regions that give rise to future tissues and organs \cite{lim2013kinetics}.
%
Images in Figure~\ref{subfig:images_unordered} show optical sections perpendicular to the long axis of the embryo.
%%%
%%% we have to somehow say that it is consistent
%%%
%
%%%
%%% ask if "visualize" is typical as active verb
%%%
The embryos were fixed and stained with chemicals that help visualize nuclei (shown in gray) as well as the spatial distributions of two different proteins (shown in green and red) \cite{chung2010microfluidic}.
%
Each of the images reveals a uniform arrangement of nuclei and a nonuniform distribution of the two proteins, which are involved in patterning of the dorsoventral (back-to-belly) axis of the embryo.

@@

%%%
%%% orientation does not read good to Y below -
%%% it is the same orientation, but a different rotation around the axis
%%% how to say this ?
%%%
Each image in Figure~\ref{subfig:images_unordered} comes from a different embryo, captured at a different rotation about the long axis and at a different and unknown point in time.
%
The goal is to register these images in a globally consistent way and order them to be representative of the underlying developmental trajectory.
%
Both of these tasks can be fully automated, and even combined.
%
We will use {\em angular synchronization} \cite{singer2011angular}, a registration algorithm that is robust to noise, to align the images in a consistent way.
%
We will use {\em diffusion maps} \cite{coifman2005geometric}, a broadly applicable technique for the discovery of low-dimensional structures in high-dimensional data sets, to temporally order our data.
%
We will also show how both steps can be combined using {\em vector diffusion maps} \cite{singer2012vector}.
%
%The paper is organized as follows.
%
%First, we use a synthetic dataset, where the correct orientation and ordering of data points is known, to formalize the tasks of image registration and temporal ordering.
%
%Then, we show that both of these tasks can be effectively combined in a single step.
%
%Following this, we return to the original set of two-dimensional images and demonstrate that our approach can readily align them and order them correctly in time.
%
%We conclude by discussing the general requirements that must be satisfied for our approach to work and compare it to other approaches and ideas in cross-sectional studies of biological dynamics.

@@

We will first demonstrate our techniques for registration and temporal ordering using a synthetic data set whose features resemble those in our imaging data set in Figure~\ref{subfig:images_unordered}.
%
This data set will allow us to illustrate the main points of the different algorithms, and will also have relatively simple dynamics which are easy to visualize.
%
We construct concentration profiles defined on a ring, and we assume that each ring is arbitrarily rotated around its center; an example is shown in Figure~\ref{subfig:1d_example}.
%
Rotation of the ring corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown in the bottom of Figure~\ref{subfig:1d_example}; the symmetry group is the group of all two-dimensional proper rotations, denoted $SO(2)$.
%
Each concentration profile is a noisy Gaussian (shown in Figure~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as  a function of time.
%
We discretize the profiles into $100$ points, so our numerical data will be $100$-dimensional vectors (the corresponding symmetry group for the discretized profiles is $\mathbb{Z}_n$, the group of integers modulo $n$).
%
Figure \ref{subfig:1d_unaligned_unordered} shows the entire data set; the concentration profiles have been stacked in an array, so that each row corresponds to a single profile.
%
Because the profiles are unregistered and unordered, the underlying dynamics (a Gaussian whose amplitude grows in time) are not readily apparent.

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig3}
\caption{(a) One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom). (b) Intensity corresponding to the profile in (a). (c) An ensemble of concentration profiles, each of the form described in (a). Each row in the array corresponds to a single profile. (d) The profiles in (c), now registered using angular synchronization. (e) The profiles in (c), now temporally ordered using diffusion maps. (f) The profiles in (c), registered and temporally ordered in a single step using vector diffusion maps.}
%\label{fig:1d_demo}
\customlabel{fig:1d_demo}{3}
\customlabel{subfig:1d_example}{\ref{fig:1d_demo}a}
\customlabel{subfig:1d_intensity}{\ref{fig:1d_demo}b}
\customlabel{subfig:1d_unaligned_unordered}{\ref{fig:1d_demo}c}
\customlabel{subfig:1d_aligned_unordered}{\ref{fig:1d_demo}d}
\customlabel{subfig:1d_aligned_ordered}{\ref{fig:1d_demo}e}
\customlabel{subfig:1d_aligned_ordered_vdm}{\ref{fig:1d_demo}f}
\end{figure}

@@

We begin by registering the set of profiles using angular synchronization\cite{singer2011angular}, a technique which is robust to noise and requires no {\em a priori} knowledge of image features.
%
Angular synchronization first aligns every possible pair of data points, i.e., it computes the rotation angles
required to make each pair have the maximum possible overlap (see {\it Materials and Methods}).
%%%
%%% so I do not like the figure with the vectors - it is correct, but people have to take the
%%% function, discretize and make it a vector in their heads, and may be "one step too much"
%%%
%
An illustration is shown in Figure~\ref{subfig:synch1};
each data point is a vector, and $\theta_{ij}$ denotes the angle of rotation needed to align vectors $i$ and $j$.
%
From this large collection of pairwise alignments, the algorithm then selects a set of rotation angles $\theta_i$, one for each data point, such that differences in rotation angles $\theta_i-\theta_j$ are {\em most consistent} (in some metric) with the pairwise alignments $\theta_{ij}$.
%
Figure~\ref{subfig:synch2} shows each vector from Figure~\ref{subfig:synch1}, now rotated so that the entire set is aligned; the computed rotation angles are also indicated.
%
Note that the differences between the rotation angles assigned to each vector are consistent with the pairwise angles needed to rotate one vector to another, i.e., $\theta_i - \theta_j \approx \theta_{ij}$.
%
To compute these alignments, the task is formulated as an optimization problem, and then relaxed to an eigenproblem.
%
The top eigenvector(s) of a matrix (which contains information about all pairwise alignments)
provides the rotation angles assigned to each data point (see {\it Materials and Methods}).
%
%Because the method considers {\em all} pairwise alignments, angular synchronization is often robust to noise.

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig4}
\caption{(a) Set of vectors, each in a different orientation. The pariwise alignment angles are indicated. (b) The vectors from (a), each rotated so that the set is globally aligned. Note that the chosen rotation angles are consistent with the pairwise alignments in (a). (c) Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of data points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see \eqref{eq:dmaps_W}), so that close data points are connected by dark (``stronger'') edges. (d) The data in (c), colored by the first (non-trivial) eigenvector from the diffusion map computational procedure, $\phi_2$. The color intensity is monotonic with the arclength, thus parameterizing the curve.  }
%\label{fig:schematics}
\customlabel{fig:schematics}{4}
\customlabel{subfig:synch1}{\ref{fig:schematics}a}
\customlabel{subfig:synch2}{\ref{fig:schematics}b}
\customlabel{subfig:dmaps_edges}{\ref{fig:schematics}c}
\customlabel{subfig:dmaps_color}{\ref{fig:schematics}d}
\end{figure}

Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles from Figure~\ref{subfig:1d_unaligned_unordered}, now aligned using angular synchronization.
%
It is reassuring that the peaks are aligned,
even though the algorithm uses no specific information about the existence or location of such peaks.
%
%We note once more that, since angular synchronization only uses pairwise rotation information,
%the optimal rotations are recovered up to a global rotation.

@@

After geometrically registering the data (concentration profiles or images), we want to order them in time to reconstruct the developmental dynamics.
%
%Our main assumption is that most of the variation in the data correlates with time.
%
We use diffusion maps \cite{coifman2005geometric}, a dimensionality reduction technique which aims to uncover low-dimensional, nonlinear structure in high-dimensional data sets, to temporally order our data.
%
Figure~\ref{subfig:dmaps_edges} shows an example of a two-dimensional data set (the ``high-dimensional" data) which lies on a one-dimensional nonlinear curve (the ``low-dimensional" manifold).
%
Figure~\ref{subfig:dmaps_color} shows the same data, now colored by the leading diffusion map coordinate.
%
The color parameterizes the one-dimensional curve that our eye can visually detect.
%
Our main assumptions are that our (one-dimensional profile, or two-dimensional image) data also lie on a one-dimensional, nonlinear curve, {\em and} that this curve is parameterized by time (e.g., in Figure~\ref{subfig:dmaps_color}, we assume that the data points are ordered from youngest to oldest along the curve).
%
Therefore, ordering data along this curve will effectively order data in time.
%
Because our data will be very high dimensional, we cannot do this by eye, and instead require an algorithm to automatically extract this one-dimensional structure.

To uncover a parameterization of this curve, we assume that data points which are close in the high-dimensional space should also be close {\em on the curve} (we would like images which look similar to be close together in our developmental trajectory).
%
%We therefore want to find a parameterization of the points, such that the points which are close in the original, high-dimensional space are {\em as close as possible} in the parameterization.
%
To emphasize this point, edges between the pairs of data points in Figure~\ref{subfig:dmaps_edges} are colored according to the distance between the data points, so that close data points have dark/strong
edges connecting them.
%
Visually, these dark edges delineate the one-dimensional curve; in contrast, the light edges are less informative
about the low-dimensional structure.
%
%We want points which are close in the high-dimensional space, i.e., connected by dark edges, to be close in our low-dimensional %parameterization.
%
Finding such a parameterization can be formulated as an optimization problem \cite{Belkin2003}, and the solution is given by the top (non-trivial) eigenvector of a Markov matrix which contains information about pairwise distances between the data points (see {\it Materials and Methods}).
%
This procedure has been named diffusion maps \cite{coifman2005geometric}.
%
%The formulation extends from curves to manifolds (i.e., we can search for an $n$-dimensional, rather than a one-dimensional, parameterization of the data, such that data points which are close in the high-dimensional space are also close in the $n$-dimensional parameterization).

Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we can use diffusion maps to order them in time.
%
We use the Euclidean distance between the (discretized in $\mathbb{R}^{100}$) aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our pairwise distances between data points.
%
Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the first (non-trivial) eigenvector, $\phi_2$.
%
As expected, the intensity of the peak grows as a function of $\phi_2$ (since we constructed these data
to have a peak growing in time).
%
%Although diffusion maps can order the data, it cannot tell us about the speed of this trajectory.
%
%Furthermore, the direction of the trajectory is also unknown; determining which end of the trajectory corresponds to the beginning of development is a task that needs to be done by hand.

@@

%We are often interested in both registering and ordering our data.
%
%Our proposed alignment methodology, angular synchronization, utilizes the eigendecompostion of the matrix $H \in \mathbb{R}^{md \times md}$, and our proposed ordering algorithm, diffusion maps, utilizes the eigendecomposition of the matrix $A \in \mathbb{R}^{m \times m}$.
%
Angular synchronization (for registration) can be combined with diffusion maps (for temporal ordering) in a single eigencomputation that allows us to {\em simultaneously} recover the alignments {\em and} the temporal ordering.
%
This technique is called vector diffusion maps \cite{singer2012vector} (see {\it Materials and Methods}).
%
It both reduces the number of required eigencomputations, and is much more robust to noise, as mistakes in alignment are also considered when ordering the data.
%
Figure \ref{subfig:1d_aligned_ordered_vdm} shows the result of simultaneously registering and ordering the data in Figure \ref{subfig:1d_unaligned_unordered} using vector diffusion maps.
%
The data has been ordered by $\psi_{1, 3}$, the first (non-trivial) vector diffusion maps coordinate.

@@

In these experiments, we are fortunate enough to be able to validate our  automatic registration and ordering using prior physical knowledge about the system.
%
For registration, it is known that the peak in the distribution of Dorsal (shown in green) specifies the ventralmost position in the embryo, and that the nuclei (shown in gray) delineate the circumference of the embryo.
%%%
%%% maybe you can also quantify this guy
%%%
%
%Based on this, images can be translated to make the centers of the embryos coincide and rotated to place the peak of the green channel at the same angular position.
%
Temporal ordering, on the other hand, can be done based on the monotonic progress of cellularization (Figure~\ref{subfig:membranes}) \cite{figard2013plasma}.%; this helps validate our vector diffusion map-based ordering.
%
The results can be visually compared through the sequences of one-dimensional profiles (similar to the format in Figure~\ref{fig:1d_demo}), shown in Figures~\ref{subfig:1d_membranes}~and~\ref{subfig:1d_vdm}.
%
The general trends are consistent:
both green  and red peaks are aligned, and the red peaks grow as a function of time (or of the leading vector diffusion map coordinate).
%
However, Figure~\ref{subfig:1d_vdm}, which was automatically registered and ordered, shows an arguably better clustering of the ``outer" peaks in the top corners of the image.
%
%These peaks are more scrambled in Figure~\ref{fig:membrane_compare}b, which was analyzed using cellularization.
%
These peaks are known to arise later in development, and so it is reassuring that they appear at the end of the (recovered) trajectory.
%
%Furthermore, the membranes stop growing towards the end of this trajectory, making the ordering of the late signals in Figure~\ref{subfig:1d_membranes} less accurate.
%
Quantitatively, the rank correlation coefficient between these two temporal orderings is 0.9149, indicating that our automatic ordering is consistent with the ordering obtained from cellularization.

@@

Given data points $x_1, \dots, x_m$, for vector diffusion maps, one first constructs the matrix $S \in \mathbb{R}^{md \times md}$, with the $i,j^{th}$ block of $S$, $S_{ij}$, defined as
\begin{equation} \label{eq:vdm_S}
	S_{ij} = A_{ij} H_{ij}
\end{equation}
%
where $A_{ij} \in \mathbb{R}$ (defined in \eqref{eq:dmaps_A}) pertains to the diffusion kernel between data points, and $H_{ij} \in \mathbb{R}^{d \times d}$ (defined in \eqref{eq:H_to_R}) pertains to the pairwise alignment between data points.
%
It is important to note that distance $d(x_i, x_j)$ used in the diffusion kernel in \eqref{eq:dmaps_W} is the distance between data points {\em after} after pairwise alignment, i.e., the minimum distance between all possible shifts of the two data points (which is obtained in \eqref{eq:opt_pairwise}).
%
In the language of symmetry groups, this distance is a metric between the orbits of the data points induced by the symmetry group.

>>>>

\dropcap{E}xperimental studies of developmental dynamics fall in two broadly defined categories: longitudinal and cross-sectional \cite{diggle2002analysis}.
%
In longitudinal studies, developmental progress is monitored over time for the same embryo.
%
In a cross-sectional study, one embryo contributes only a single snapshot of a chemical or morphological process along its developmental trajectory, and the developmental dynamics must be reconstructed from multiple snapshots of different embryos.
%
Both of these sampling schemes have their advantages and limitations, and both are extensively used by developmental biologists.
%
Here we focus on cross-sectional studies, which have a time-honored history and still present the only option for most organisms.
%
In a typical cross-sectional study, a group of developing embryos is fixed using a procedure that arrests their development and stained with chemicals that help visualize a handful of cellular processes.
%
Fixed embryos are then imaged using any given number of microscopy techniques.
%
Recent advances in large-scale physical manipulation and imaging of embryos have produced rapidly increasing volumes of cross-sectional data, in which every embryo is observed at a different geometric orientation and developmental time point.
%
Importantly, the ``age'' of any given embryo arrested in its development is not known to high accuracy.
%
In general, it is only known that a collection of embryos belongs to a certain time window.
%
In order to recover the developmental dynamics from such image datasets, snapshots of different embryos must be spatially aligned or registered to factor out the relevant symmetries (i.e., translations and rotations), and then ordered in time.
%
We present a general computational framework which greatly accelerates -and can combine- both of these tasks.

Temporal ordering and registration of images is straightforward when the number of images is small and differences between them are visually apparent.
%
As an example, Figure~\ref{fig:fish} shows a caricature of fish development, combining the processes of growth and color patterning.
%
In this case, temporal ordering can be accomplished by arranging the fish by size, which is monotonic with the developmental progress.
%
Image registration is based on obvious morphological landmarks, such as the position of head and fins.
%
On the other hand, real data poses nontrivial challenges, due to the number of images, measurement noise and variability, and the subtlety of the developmental changes.

To illustrate the issues associated with temporal ordering and registration of images, we use a data set from an imaging study of signaling protein expression in the early fruit fly embryo, one of the leading model systems for studies of developmental dynamics.
%
This dataset comprises $49$ fluorescence microscopy images of different embryos fixed during the $3^{rd}$ hour of their development.
%
\begin{figure}[t]
\includegraphics[width=8.4cm]{fig1}
\caption{(a) Fish, each in a different orientation and a different stage of development. (b) Fish, now rotationally registered and temporally ordered. For this caricature, the registration and ordering is easy to do ``by hand" because the data set is small and the developmental changes are easy to recognize.}
%\label{fig:fish}
\customlabel{fig:fish}{1}
\customlabel{subfig:fish_unordered}{\ref{fig:fish}a}
\customlabel{subfig:fish_ordered}{\ref{fig:fish}b}
\end{figure}
%
The fruit fly embryo is shaped like a rice grain, approximately $1/2$~mm long.
%
By end of the $2^{nd}$ hour of development, sequential divisions of a fertilized egg generate a system where $\sim 6,000$ nuclei are uniformly arranged under the common plasma membrane.
%
During the $3^{rd}$ hour, the embryo undergoes the process of cellularization, whereby membranes that enclose nuclei into different cells grow in thickness (Figure~\ref{subfig:membranes}).
%
In conjunction with these morphological changes, concentration profiles of multiple regulatory molecules form inside the embryo, subdividing it into regions that give rise to future tissues and organs \cite{lim2013kinetics}.
%
Images in Figure~\ref{subfig:images_unordered} show optical sections at a fixed depth, perpendicular to the long axis of the embryo.
%%%
%%% we have to somehow say that it is consistent
%%%
%
%%%
%%% ask if "visualize" is typical as active verb
%%%
The embryos were fixed and stained with chemicals that help visualize nuclei (shown in gray) as well as the spatial distributions of two different proteins (shown in green and red) \cite{chung2010microfluidic}.
%
Each of the images reveals a uniform arrangement of nuclei and a nonuniform distribution of the two proteins, which are involved in patterning of the dorsoventral (back-to-belly) axis of the embryo.

@@

%%%
%%% orientation does not read good to Y below -
%%% it is the same orientation, but a different rotation around the axis
%%% how to say this ?
%%%
Each image in Figure~\ref{subfig:images_unordered} comes from a different embryo, captured at a different rotation about the long axis and at a different and unknown point in time.
%
The goal is to register these images in a globally consistent way, and order them to be representative of the underlying developmental trajectory.
%
Both of these tasks can be fully automated, and even combined.
%
We will use {\em angular synchronization} \cite{singer2011angular}, a registration algorithm that is robust to noise, to align the images in a consistent way.
%
We will use {\em diffusion maps} \cite{coifman2005geometric}, a broadly applicable technique for the discovery of low-dimensional structures in high-dimensional data sets, to temporally order our data.
%
We will also show how both steps can be combined using {\em vector diffusion maps} \cite{singer2012vector}.
%
%The paper is organized as follows.
%
%First, we use a synthetic dataset, where the correct orientation and ordering of data points is known, to formalize the tasks of image registration and temporal ordering.
%
%Then, we show that both of these tasks can be effectively combined in a single step.
%
%Following this, we return to the original set of two-dimensional images and demonstrate that our approach can readily align them and order them correctly in time.
%
%We conclude by discussing the general requirements that must be satisfied for our approach to work and compare it to other approaches and ideas in cross-sectional studies of biological dynamics.

@@

We will first demonstrate our techniques for registration and temporal ordering using a synthetic data set whose features resemble the image features in Figure~\ref{subfig:images_unordered}.
%
The relatively simple dynamics of this data set will allow us to easily visualize and illustrate the main points of the different algorithms.
%
We construct concentration profiles defined on a ring, and we assume that each ring is arbitrarily rotated around its center; an example is shown in Figure~\ref{subfig:1d_example}.
%
Rotation of the ring corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown in the bottom of Figure~\ref{subfig:1d_example} (the symmetry group is $SO(2)$, the group of all two-dimensional proper rotations).
%
Each concentration profile is a noisy Gaussian (shown in Figure~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as  a function of time.
%
We discretize the profiles into $100$ points, so our numerical data will be $100$-dimensional vectors (the corresponding symmetry group for the discretized profiles is $\mathbb{Z}_n$, the group of integers modulo $n$).
%
Figure \ref{subfig:1d_unaligned_unordered} shows the entire data set; the concentration profiles have been stacked in an array, so that each row corresponds to a single profile.
%
Because the profiles are unregistered and unordered, the underlying dynamics (a Gaussian whose amplitude grows in time) are not readily apparent.

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig3}
\caption{(a) One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom). (b) Intensity corresponding to the profile in (a). (c) An ensemble of concentration profiles, each of the form described in (a). Each row in the array corresponds to a single profile. (d) The profiles in (c), now registered using angular synchronization. (e) The profiles in (d), now temporally ordered using diffusion maps. (f) The profiles in (c), registered and temporally ordered in a single step using vector diffusion maps.}
%\label{fig:1d_demo}
\customlabel{fig:1d_demo}{3}
\customlabel{subfig:1d_example}{\ref{fig:1d_demo}a}
\customlabel{subfig:1d_intensity}{\ref{fig:1d_demo}b}
\customlabel{subfig:1d_unaligned_unordered}{\ref{fig:1d_demo}c}
\customlabel{subfig:1d_aligned_unordered}{\ref{fig:1d_demo}d}
\customlabel{subfig:1d_aligned_ordered}{\ref{fig:1d_demo}e}
\customlabel{subfig:1d_aligned_ordered_vdm}{\ref{fig:1d_demo}f}
\end{figure}

@@

We begin by registering the set of profiles using angular synchronization\cite{singer2011angular}, a technique which is robust to noise and requires no {\em a priori} knowledge of image features.
%
Angular synchronization first aligns every possible pair of data points, i.e., it computes the rotation angles
which make each pair have the maximum possible overlap (see {\it Materials and Methods}).
%%%
%%% so I do not like the figure with the vectors - it is correct, but people have to take the
%%% function, discretize and make it a vector in their heads, and may be "one step too much"
%%%
%
An illustration is shown in Figure~\ref{subfig:synch1};
each data point is a vector, and $\theta_{ij}$ denotes the angle of rotation needed to align vectors $i$ and $j$.
%
From this large collection of pairwise alignments, the algorithm then selects a set of rotation angles, $\theta_i$ (one for each data point), such that differences in rotation angles, $\theta_i-\theta_j$, are {\em most consistent} (in some metric) with the pairwise alignments $\theta_{ij}$.
%
Figure~\ref{subfig:synch2} shows each vector from Figure~\ref{subfig:synch1}, now rotated so that the entire set is aligned; the computed rotation angles are also indicated.
%
Note that the differences between the rotation angles assigned to each vector are consistent with the pairwise angles needed to rotate one vector to another, i.e., $\theta_i - \theta_j \approx \theta_{ij}$.
%
The problem of finding these alignments is formulated as an optimization problem, and then relaxed to an eigenproblem.
%
The top eigenvector(s) of a matrix (which contains information about all pairwise alignments)
provides the rotation angles assigned to each data point (see {\it Materials and Methods}).
%
%Because the method considers {\em all} pairwise alignments, angular synchronization is often robust to noise.
%
\begin{figure}[t]
\includegraphics[width=8.4cm]{fig4}
\caption{(a) Set of vectors, each in a different orientation. The pariwise alignment angles are indicated. (b) The vectors from (a), each rotated so that the set is globally aligned. Note that the chosen rotation angles are consistent with the pairwise alignments in (a). (c) Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of data points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see \eqref{eq:dmaps_W}), so that close data points are connected by dark (``stronger'') edges. (d) The data in (c), colored by the first (non-trivial) eigenvector from the diffusion map computational procedure, $\phi_2$. The color intensity is monotonic with the arclength, thus parameterizing the curve.  }
%\label{fig:schematics}
\customlabel{fig:schematics}{4}
\customlabel{subfig:synch1}{\ref{fig:schematics}a}
\customlabel{subfig:synch2}{\ref{fig:schematics}b}
\customlabel{subfig:dmaps_edges}{\ref{fig:schematics}c}
\customlabel{subfig:dmaps_color}{\ref{fig:schematics}d}
\end{figure}
%
Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles from Figure~\ref{subfig:1d_unaligned_unordered}, now aligned using angular synchronization.
%
It is reassuring that the peaks are aligned,
even though the algorithm uses no specific information about the existence or location of such peaks.
%
%We note once more that, since angular synchronization only uses pairwise rotation information,
%the optimal rotations are recovered up to a global rotation.

@@

After geometrically registering the data (concentration profiles or images), we want to order them in time to reconstruct the developmental dynamics.
%
%Our main assumption is that most of the variation in the data correlates with time.
%
Our main assumptions are that the high-dimensional data (images have one dimension per color pixel intensity, and concentration profiles have one dimension per discretization point) lie on a one-dimensional, nonlinear curve, {\em and} that this curve is parameterized by time.
%
Figure~\ref{subfig:dmaps_edges} shows an schematic of a two-dimensional data set (the ``high-dimensional" data) which lies on a one-dimensional nonlinear curve (the ``low-dimensional" manifold).
%
We assume that the data points are ordered from youngest to oldest along the curve.
%
Because our data are very high dimensional, we cannot extract this curve by eye.
%
We instead use diffusion maps \cite{coifman2005geometric}, a dimensionality reduction technique which aims to uncover low-dimensional, nonlinear structure in high-dimensional data sets, to automatically extract this one-dimensional curve.
%
Figure~\ref{subfig:dmaps_color} shows the same data, now colored by the leading diffusion map coordinate.
%
The color parameterizes the one-dimensional curve that our eye can visually detect, and ordering the data by the leading diffusion map coordinate will effectively order data in time.

To uncover the diffusion maps parameterization, we assume that data points which are close in the high-dimensional space should also be close {\em on the curve} (we would like images which look similar to be close together in our developmental trajectory).
%
%We therefore want to find a parameterization of the points, such that the points which are close in the original, high-dimensional space are {\em as close as possible} in the parameterization.
%
To emphasize this point, edges between the pairs of data points in Figure~\ref{subfig:dmaps_edges} are colored according to the distance between the data points, so that pairs of points which are close have dark/strong edges connecting them.
%
Visually, the dark edges delineate the one-dimensional curve, whereas the light edges are less informative about the low-dimensional structure.
%
%We want points which are close in the high-dimensional space, i.e., connected by dark edges, to be close in our low-dimensional %parameterization.
%
Finding such a parameterization can be formulated as an optimization problem \cite{Belkin2003}, and the solution is given by the top (non-trivial) eigenvector of a Markov matrix which contains information about pairwise distances between the data points (see {\it Materials and Methods}).
%
%This procedure has been named diffusion maps \cite{coifman2005geometric}.
%
%The formulation extends from curves to manifolds (i.e., we can search for an $n$-dimensional, rather than a one-dimensional, parameterization of the data, such that data points which are close in the high-dimensional space are also close in the $n$-dimensional parameterization).

Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we use diffusion maps to order them in time.
%
We use the Euclidean distance between the (discretized in $\mathbb{R}^{100}$) aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our pairwise distances between data points.
%
Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the leading diffusion maps coordinate, $\phi_2$.
%
As expected, the intensity of the peak grows as a function of $\phi_2$ (since we constructed these data
to have a peak growing in time).
%
%Although diffusion maps can order the data, it cannot tell us about the speed of this trajectory.
%
%Furthermore, the direction of the trajectory is also unknown; determining which end of the trajectory corresponds to the beginning of development is a task that needs to be done by hand.

@@

%We are often interested in both registering and ordering our data.
%
%Our proposed alignment methodology, angular synchronization, utilizes the eigendecompostion of the matrix $H \in \mathbb{R}^{md \times md}$, and our proposed ordering algorithm, diffusion maps, utilizes the eigendecomposition of the matrix $A \in \mathbb{R}^{m \times m}$.
%
Angular synchronization (for registration) can be combined with diffusion maps (for temporal ordering) in a single eigencomputation that allows us to {\em simultaneously} recover the alignments {\em and} the temporal ordering.
%
This technique is called vector diffusion maps \cite{singer2012vector} (see {\it Materials and Methods}).
%
It both reduces the number of required eigencomputations, and is much more robust to noise, as mistakes in alignment are also considered when ordering the data.
%
Figure \ref{subfig:1d_aligned_ordered_vdm} shows the result of simultaneously registering and ordering the data in Figure \ref{subfig:1d_unaligned_unordered} using vector diffusion maps.
%
The data has been ordered by $\psi_{1, 3}$, the leading vector diffusion maps coordinate.

@@

In these experiments, we are fortunate enough to be able to validate our  automatic registration and ordering using prior physical knowledge about the system.
%
For registration, it is known that the peak in the distribution of Dorsal (shown in green) specifies the ventralmost position in the embryo, and that the nuclei (shown in gray) delineate the circumference of the embryo.
%%%
%%% maybe you can also quantify this guy
%%%
%
%Based on this, images can be translated to make the centers of the embryos coincide and rotated to place the peak of the green channel at the same angular position.
%
Temporal ordering, on the other hand, can be done based on the monotonic progress of cellularization (Figure~\ref{subfig:membranes}) \cite{figard2013plasma}.%; this helps validate our vector diffusion map-based ordering.
%
The two methods for registration and ordering can be visually compared through the sequences of one-dimensional profiles extracted from the circumference of the embryos (similar to the format in Figure~\ref{fig:1d_demo}), shown in Figures~\ref{subfig:1d_membranes}~and~\ref{subfig:1d_vdm}.
%
The general trends are consistent:
both green  and red peaks are aligned, and the red peaks grow as a function of time/leading vector diffusion map coordinate.
%
However, Figure~\ref{subfig:1d_vdm}, which was automatically registered and ordered, shows an arguably better clustering of the ``outer" peaks in the top corners of the image.
%
%These peaks are more scrambled in Figure~\ref{fig:membrane_compare}b, which was analyzed using cellularization.
%
These peaks are known to arise later in development, and so it is reassuring that they appear at the end of the (recovered) trajectory.
%
%Furthermore, the membranes stop growing towards the end of this trajectory, making the ordering of the late signals in Figure~\ref{subfig:1d_membranes} less accurate.
%
Quantitatively, the rank correlation coefficient between these two temporal orderings is 0.9158, indicating that our automatic ordering is consistent with the ordering obtained from cellularization.

@@

In vector diffusion maps, given data points $x_1, \dots, x_m$, one first constructs the matrix $S \in \mathbb{R}^{md \times md}$, with the $i,j^{th}$ block of $S$, $S_{ij}$, defined as
\begin{equation} \label{eq:vdm_S}
	S_{ij} = A_{ij} H_{ij}
\end{equation}
%
where $A_{ij} \in \mathbb{R}$ (defined in \eqref{eq:dmaps_A}) pertains to the diffusion kernel between data points, and $H_{ij} \in \mathbb{R}^{d \times d}$ (defined in \eqref{eq:H_to_R}) pertains to the pairwise alignment between data points.
%
It is important to note that distance $d(x_i, x_j)$ used in the diffusion kernel in \eqref{eq:dmaps_W} is the distance between data points {\em after} after pairwise alignment, i.e., the minimum distance between all possible shifts of the two data points (which is obtained in \eqref{eq:opt_pairwise}).
%
In the language of symmetry groups, this distance is a metric between the orbits induced by the symmetry group.
